---
title: "Danish Kings"
author: "Helena Hymøller Jensen"
date: "2026-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator: semicolon 

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
library(tidyverse)
#FILL IN THE CODE BELOW and review the outputs
kings1 <- read.csv("data/danish_kings.csv")

kings2 <- read_csv("data/danish_kings.csv")

kings3 <- read.csv2("data/danish_kings.csv")

kings4 <- read_csv2( "data/danish_kings.csv")
```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object

#read.csv() and read.csv2() are base R functions

#read_csv() and reac_csv2() are tidyverse functions; for the tidyverse functions to work the tidyverse packages must be activated

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.

#output: [1] "spec_tbl_df" "tbl_df"      "tbl"         "data.frame"

#meaning the object is a tibble, which essentially is what was expected

3. How many columns does the object have when created with these different functions?

#the output from running the code display that the object has 10 columns

4. Show the dataset so that we can see how R interprets each column

#the tidyverse function glimpse() interprets each column, by displaying the structure, the column names, and data types; it also displays the number of rows and columns

#the base R function str() can be used as well since it displays the data type, which can determine whether the data type should be converted

#the tidyverse function head() 
```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

library(tidyverse)

#the dataset contains semicolons as separators, which makes the code read_csv2() the faultless function
kings <- read_csv2("data/danish_kings.csv")

#identifying the type of object
class(kings)

#displaying the first couple of rows of the dataset
head(kings)

#identifying how the column are interpreted
glimpse(kings)

```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
# YOUR CODE

#coercing the data to numeric data; focusing on the columns reign_start and reign_end
#the columns contain missing values; for that reason the function na.rm=NULL is added
kings <- kings %>% 
  mutate(
    reign_start=as.numeric(reign_start,na.rm=NULL),
    reign_end=as.numeric(reign_end)
  )

#creating a new column by calculating values from two already existing columns
kings <- kings %>% 
  mutate(duration=reign_end-reign_start,na.rm=TRUE)

glimpse(kings)

```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r}
# YOUR CODE

#calculating the average duration of reign by using the duration column as a vector; the argument na.rm=NULL ensures that missing values are ignored in the calculation
mean(kings$duration,na.rm=TRUE)

#calculating the average duration of reign using the datatset as a tibble
kings %>% 
  summarize(avg_duration=mean(duration,na.rm=TRUE))

#both methods - the vector approach using base R and the tibble approach using tidyverse - calculate the exact same average duration of reign

```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r}
# YOUR CODE

#saving the average duration of reign as an object
avg_duration <- mean(kings$duration, na.rm = TRUE)

#filtering the monarchs reigning longer than average; creating a new object
long_reigning_kings <- kings %>% 
  filter(duration>avg_duration)

#creating a tibble counting the number of long reigning kings
long_reigning_kings %>% 
  count()

```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
# YOUR CODE

#sorting the kings by duration in descending order
longest_ruling_kings <- kings %>% 
  arrange(desc(duration))

#selecting the three longest ruling monarchs by essentially extracting rows one, two, and three
three_longest_ruling <- longest_ruling_kings %>% 
  slice(1:3)

#calulating the total number of days by creating a new column
three_longest_ruling <- three_longest_ruling %>% 
  mutate(days=duration*365)

#calculating the total number of days considering leap years
three_longest_ruling <- three_longest_ruling %>% 
  mutate(days=duration*365.25)

#displaying the results
three_longest_ruling
view(three_longest_ruling)

```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# YOUR CODE

#creating a new column; midpoints contributes to a more balanced representation reflecting how the duration of reigns changed over time
kings <- kings %>% 
  mutate(midyear=reign_start+(reign_end-reign_start)/2)

#the following code sets up a plot with midyear as x-values and duration as y-values
#each monarch is represented as the midpoint of their reign, but adding a smoothed trend line showcases a general trend in the data
#stylistic wishes creates the need for including the codes of labs() adding in a title and labels for the axis; themes for font and size are added in as well
ggplot(kings,aes(x=midyear,y=duration))+
  geom_point()+
  geom_smooth()+
  labs(
    title = "The Duration of Reigns Throughout Time",
    x="Midpoint of Reign",
    y="Duration of Reign"
  )+
  theme_bw()+
  theme(text = element_text(size = 14))

##the overall trend, showcased in the plot, indicates that the duration of the reign overall tended to lengthen over time; many structural factors led to this development such as political stability, improved governance and technological advancements

```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


